---
title: "basic_tutorial"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{basic_tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r echo=FALSE}
# set random seed
set.seed(1)

# load the SIMPLEGEN package
library(SIMPLEGEN)
```

This tutorial goes through the basic process of creating a SIMPLEGEN project and running the inbuilt epidemiological simulator. 

### Running a basic simulation

SIMPLEGEN works with projects, which are essentially just lists that hold all inputs and outputs in a single convenient place. We start by creating a new project and specifying the parameters of the epidemiological model - see `?define_epi_params` for a complete list of model parameters and default values.

```{r}
myproj <- simplegen_project()
myproj <- define_epi_params(myproj,
                            H = 1e3,
                            M = 1e4)
```

Alternatively, if you are familiar with the pipe operator `%>%` from the `dplyr` package then you can use this to chain together multiple commands:

```{r eval=FALSE}
library(dplyr)
myproj <- simplegen_project() %>%
  define_epi_params()
```

Taking a quick look at our project we will see a summary of the current setup:

```{r}
myproj
```

A good way of sanity-checking our model before running simulations is to use some of the inbuilt plotting functions to explore our assumed distributions. For example, we can look at the assumed distribution of time in the acute state as follows:

```{r}
plot_epi_distribution(myproj, name = "duration_acute")
```

Once we are happy with our assumptions we are ready to run the epidemiological simulation. Note that the argument `pb_markdown = TRUE` is used here just to avoid cluttering this tutorial with too much output, but you should run the model without this argument.

```{r}
myproj <- sim_epi(myproj, pb_markdown = TRUE)
```

Looking at the project again we now see that we have some results:

```{r}
myproj
```

There are custom functions within SIMPLEGEN for exploring model output (see below), or we can dig into the output manually using the `$` symbol, for example `myproj$epi_output$daily_values`.


### Plotting results

One of the most basic plots that we might be interested in is the daily number of hosts in each state. We can specify the states that we want to plot as an input vector:

```{r}
plot_daily_states(myproj, states = c("A", "C", "P"))
```

This is also true of mosquito states, which are suffixed with "v":

```{r}
plot_daily_states(myproj, states = c("Ev", "Iv"))
```

In addition to storing daily counts in each state, the `sim_epi()` function takes a slice of the population at various times, allowing us to plot age-distributions at these points in time. The sampling times are specified by the argument `output_age_times`, which defaults to the final timepoint only. Hence, we can plot:

```{r}
plot_age_states(myproj, state = "C")
```


### Multiple demes and migration

SIMPLEGEN allows for multiple sub-populations (demes), which can be perfectly isolated or connected by migration. First, we specify the human and mosquito population sizes in each deme by using a vector of values for `H`, `M`, and `seed_infections`. Second, we specify a migration matrix giving the daily probability of each human moving from the source deme (rows) to the destination deme (columns).

```{r}
# define parameters over multiple demes
n_demes <- 3
H <- rep(1e3, n_demes)
seed_infections <- c(1e2, 0, 0)
M <- c(1e3, 2e3, 3e3)

# define a migration matrix
m <- 0.01
mig_mat <- matrix(m / n_demes, n_demes, n_demes) + diag(1 - m, n_demes)

# load new paramters into the model
myproj <- define_epi_params(myproj,
                            H = H,
                            seed_infections = seed_infections,
                            M = M,
                            mig_mat = mig_mat)
```

We then run the simulation as normal:

```{r}
myproj <- sim_epi(myproj, pb_markdown = TRUE)
```

Now when plotting results we can use the `deme` argument to specify which deme to plot:

```{r}
plot_daily_states(myproj, states = c("A", "C", "P"), deme = 1)
plot_daily_states(myproj, states = c("A", "C", "P"), deme = 2)
plot_daily_states(myproj, states = c("A", "C", "P"), deme = 3)
```

In the example above there are three demes, with only the first containing infected hosts initially. As time progresses we  see infections seed into the other two demes via migration.

