---
title: "Installation"
author: "Bob Verity"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Installation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

### Installing and loading SIMPLEGEN

SIMPLEGEN uses the [cpp11](https://cpp11.r-lib.org/articles/cpp11.html) package to run C++ code from within R, meaning you will need a C++ compiler. To get one:

    - On Windows, install [Rtools](https://cran.r-project.org/bin/windows/Rtools/)
    - On Mac, install Xcode from the app store.
    - On Linux, sudo apt-get install r-base-dev or similar.

Next, in R, ensure that you have the [devtools](https://www.rstudio.com/products/rpackages/devtools/) package installed by running

```{r, eval = FALSE}
find.package("devtools")
```

If you already have devtools you should see a link to the package's installation, or if not you will see an error. You can install devtools by running

```{r, eval = FALSE}
install.packages("devtools", repos = 'http://cran.us.r-project.org')
```

Now you can install the SIMPLEGEN package directly from GitHub by running

```{r, eval = FALSE}
devtools::install_github("mrc-ide/SIMPLEGEN")
```

Assuming everything installed correctly, load the package using

```{r}
library(SIMPLEGEN)
```

You can test that the package is loaded and working by running the following command, which should produce the following output:

```{r}
check_SIMPLEGEN_loaded()
```

If you have any problems installing, please [raise an issue](https://github.com/mrc-ide/SIMPLEGEN/issues) on github.
